  
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and Processing Data

Unzip and load CSV file 

```{r}

library(dplyr)
library(ggplot2)

unzip(zipfile="./data/activity.zip")
activity <- read.csv("activity.csv")

```

View summary and structure of the data to see if transformations are needed

```{r}

summary(activity)
str(activity)

```

Convert date from a factor to date

```{r}
activity$date <- as.Date(activity$date)
```


## What is mean total number of steps taken per day?

Calculate steps per day

```{r}
steps.day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
head(steps.day)
```

Create histogram 

```{r}
ggplot(steps.day, aes(x = steps)) +
    geom_histogram(fill = "green4", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

Calculate mean and median total number of steps taken per day

```{r}
(mean <- mean(steps.day$step, na.rm = TRUE))
(median <- median(steps.day$step, na.rm = TRUE))

```

mean steps taken per day: **`r mean`**, median: **`r median`**

## What is the average daily activity pattern?

calculate average daily activity

```{r}
steps.mean <- aggregate(steps ~ interval, data = activity, FUN = "mean", na.rm = TRUE)

```

create time series plot

```{r}
ggplot(steps.mean, aes(x=interval, y=steps)) +
  geom_line(color="grey57", size=1) + 
  labs(title = "Average Daily Steps", y = "Avg Steps per Day", x = "Interval")

```

Calculate the interval with the max number of steps

```{r}
(max.steps <- max(steps.mean$steps))
(max.interval <- steps.mean$interval[which(steps.mean$steps == max.steps)])
```

The max number of steps was **`r max.steps`**, which was interval **`r max.interval`**

## Imputing missing values

Calculate the total number of missing values
```{r}
(missing <- sum(is.na(activity)))
```

Number of missing values: **`r missing`**

fill in missing values with interval mean across days and create new data set 

```{r}
fill.data <- activity %>%
  mutate(
    steps = case_when(
      is.na(steps) ~ steps.mean$steps[match(activity$interval, steps.mean$interval)],      
      TRUE ~ as.numeric(steps)
    ))
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r}
steps.day2 <- aggregate(steps ~ date, data = fill.data, FUN = sum, na.rm = TRUE)

ggplot(steps.day2, aes(x = steps)) +
    geom_histogram(fill = "green4", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

Calculate mean and median

```{r}
(mean <- mean(steps.day2$step, na.rm = TRUE))
(median <- median(steps.day2$step, na.rm = TRUE))

```

mean: **`r mean`**, median: **`r median`**

Values are very similar, the mean in the same but median is slightly higher when NAs are filled in with interval mean 

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
steps.days <- fill.data %>%
  mutate(weekday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday" ))
```

Create a factor variable

```{r}
steps.days$weekday <- as.factor(steps.days$weekday)
str(steps.days)
```

Create a time series plot, first grouping by weekday

```{r}
interval.weekday <- aggregate(steps ~ interval + weekday, data = steps.days, FUN = "mean", na.rm = TRUE)
head(interval.weekday)
```

Create time series plot 

```{r}
ggplot(interval.weekday, aes(x = interval, y = steps, col = weekday)) + 
  geom_line() +
  facet_grid(rows = interval.weekday$weekday) + 
  labs(x = "Interval", 
       y = "Number of Steps", 
       title = "Average steps by time of day")
```



